---
title: "Misspecified"
author: "Soham Ghosh"
date: "2023-06-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(mvtnorm)
library(tmvtnorm)
library(msm)
library(pscl)
library(ars)
 

# parallel
library(foreach)
library(doParallel)


setwd("/Users/sohamghosh/Desktop/factormodel")

```

## Parameter and Data Generation
Below, we generate the true parameters. We take $\Omega = I_m$ and randomly generate $\beta$ column by column. Then, the data $y$ is sampled from a MVN distribution with mean 0 and covariance matrix $\Sigma = \Omega + \beta \beta'$.

```{r, echo=FALSE}
# Function that generates data for factor model
# n - No of samples in data
# d - Dimension of the data
# q0 - True no of factor
# flag - Deciding homoscedastic or heteroscedastic error 
# GenerateData <- function(n, d, q0, flag = 1) {
#   if (flag == 0) {
#     psi = rep(1, d)
#   }
#   if (flag == 1) {
#     psi = (10:(10+d-1)/4)
#   }
#   
#   # Generating loading matrix
#   W_true = matrix(rnorm(d*q0, 0, 1), nrow = d, ncol = q0)
#   
#   # Generating data matrix
#   y = matrix(0, nrow = n, ncol = d)
#   for(i in 1:n){
#     f = rnorm(q0, mean = 0, sd = 1)
#     y[i,] = rmvnorm(n = 1, mean = W_true %*% f, sigma = diag(psi))
#   }
#   
#   return(y)
# }

# Function that generates data for factor model
# n - No of samples in data
# d - Dimension of the data
# q0 - True no of factor
# flag - Deciding homoscedastic or heteroscedastic error 
GenerateData <- function(n, d, q0, flag = 1) {
  if (flag == 0) {
    psi = rep(0.1, d)
  }
  if (flag == 1) {
    psi = sample(round(seq(from = 0.05, to = 0.1, length.out = d), 3), size = d)
  }
  
  # Generating loading matrix
  #W_true = matrix(rnorm(d*q0, 0, 1), nrow = d, ncol = q0)
  
  lam = sample(x = 10:14, size = q0, replace = TRUE)
  D = matrix(0, nrow = d, ncol = q0)
  diag(D) = lam
  #lam = sample(x = 10:15, size = d, replace = TRUE)
  U = pracma::randortho(d, type = "orthonormal")
  V = pracma::randortho(q0, type = "orthonormal")
  W_true = U %*% D %*% V[, 1:q0]
  
  # Generating data matrix
  y = matrix(0, nrow = n, ncol = d)
  for(i in 1:n){
    f = rnorm(q0, mean = 0, sd = 1)
    y[i,] = rmvnorm(n = 1, mean = W_true %*% f, sigma = diag(psi))
  }
  
  return(y)
}

```

```{r, echo=FALSE}
# Function that do Gibbs sampling for factor model
# y - Data matrix
# k - No of factor assumed
# beta0 - Initial factor loading matrix

factGibbsMod <- function(y, k, beta0, Sigma0, nu, C0, s, No.Iter, burnin, thin) {
  
  # Getting dimension information
  T <- nrow(y)
  m <- ncol(y)


  # Take care of the case when k = 0
  if (k == 0) {
    output <- list()
    output$beta <- NA
    output$F <- NA
    Sigma.temp <- numeric() # a draw from the posterior
    for (j in 1:m) {
      Sigma.temp[j] <- rigamma(
        n = 1, alpha = (nu + T) / 2,
        beta = (sum(y[, j]^2) + nu * s^2) / 2
      )
    }
    output$Sigma[[1]] <- Sigma.temp # split out a posterior draw for Sigma
    return(output)
  }


  ######
  # list objects for the simulated parameters initialized

  beta <- vector(mode = "list", length = No.Iter / thin)
  Sigma <- vector(mode = "list", length = No.Iter / thin)
  ##############################


  # set the current values of the parameters
  beta.cur <- beta0
  Sigma.cur <- Sigma0


  for (iter in 1:(No.Iter + burnin)) {
    # update F
    F.temp <- rmvnorm(
      n = T,
      mean = rep(0, k),
      sigma = solve(diag(k) + t(beta.cur) %*% diag(1 / Sigma.cur) %*% beta.cur)
    )


    F.temp <- F.temp + t(solve(diag(k) + t(beta.cur) %*% diag(1 / Sigma.cur) %*% beta.cur) %*% t(beta.cur) %*% diag(1 / Sigma.cur) %*% t(y))

    F.cur <- F.temp

    # update Sigma
    Sigma.temp <- numeric(m)

    for (i in 1:m) {
      if (i <= k) {
        Sigma.temp[i] <- rigamma(
          n = 1, alpha = (nu + T) / 2,
          beta = (nu * s^2 + crossprod(y[, i] - F.cur[, 1:i] %*% as.matrix(beta.cur[i, 1:i]))) / 2
        )
      } else {
        Sigma.temp[i] <- rigamma(
          n = 1, alpha = (nu + T) / 2,
          beta = (nu * s^2 + crossprod(y[, i] - F.cur %*% as.matrix(beta.cur[i, ]))) / 2
        )
      }
    }

    # update the Sigma list when necessary
    if (iter > burnin && (iter - burnin) %% thin == 1 && thin != 1) {
      Sigma[[(iter - burnin) %/% thin + 1]] <- Sigma.temp
    } else if (thin == 1 && (iter - burnin) > 0) {
      Sigma[[iter - burnin]] <- Sigma.temp
    }

    Sigma.cur <- Sigma.temp


    # update beta

    beta.temp <- matrix(0, nr = m, nc = k)

    for (i in 1:m) {
      if (i <= k) {
        Ci <- solve(1 / C0 * diag(i) + 1 / Sigma.cur[i] * crossprod(F.cur[, 1:i]))
        mi <- Ci %*% (1 / Sigma.cur[i] * t(F.cur[, 1:i]) %*% y[, i])

        if (k != 1) {
          beta.temp[i, i] <- ars(1, ldkernel,
            ldkernelprima,
            x = c(0.1, 500, 1000), m = length(c(0.1, 500, 1000)), lb = TRUE, ub = FALSE,
            xlb = 0, k = k, i = i, mu = mi[i], c = Ci[i, i]
          )
        } else {
          # to tackle the instability of ars when k = 1
          beta.temp[i, i] <- rtnorm(n = 1, mean = mi[i], sd = sqrt(Ci[i, i]), lower = 0)
        }

        if (i != 1) {
          beta.temp[i, 1:(i - 1)] <-
            rmvnorm(1,
              mean = mi[1:(i - 1)] + (1 / Ci[i, i]) * (beta.temp[i, i] - mi[i]) * Ci[1:(i - 1), i],
              sigma = Ci[1:(i - 1), 1:(i - 1)] - 1 / Ci[i, i] * tcrossprod(Ci[1:(i - 1), i], Ci[i, 1:(i - 1)])
            )
        }
      } else {
        Ci <- solve(1 / C0 * diag(k) + 1 / Sigma.cur[i] * crossprod(F.cur))
        mi <- Ci %*% (1 / Sigma.cur[i] * t(F.cur) %*% y[, i])
        beta.temp[i, 1:k] <- rmvnorm(n = 1, mean = as.numeric(mi), sigma = Ci)
      }
    }

    beta.cur <- beta.temp


    if (iter > burnin && (iter - burnin) %% thin == 1 && thin != 1) {
      beta[[(iter - burnin) %/% thin + 1]] <- beta.temp
    } else if (thin == 1 && (iter - burnin) > 0) {
      beta[[iter - burnin]] <- beta.temp
    }


    # if(iter%%1000 == 0 )
    # cat("You are done with the ",iter, "th iterations when k is ", k,
    #       append= T,
    #     sep = "\n"
    #       )
  }

  output <- list()
  output$beta <- beta
  output$Sigma <- Sigma

  return(output)
}

#############
ldkernel <- function(x, k, i, mu, c){
  return((k - i)*log(x) - 0.5* (x - mu)^2/c)
}

#############
ldkernelprima <- function(x, k, i, mu, c){
  return((k - i)/x - (x- mu)/c )
  
}

```

```{r, echo=FALSE}
# Function that initialize parameters for Gibbs sampling
# d - Dimension of the data
# q - Assumed no of factor
InitParam <- function(d, q) {
  # init value for beta0 used in Gibbs sampler
  beta_init = matrix(0, nr = d, nc = q) 
  for(i in 1:q) {
      if(i == 1){
          beta_init[1, 1] = rtnorm(n = 1, mean = 0, sd =1, lower= 0)
      } else {
          lower.vec = c(rep(-Inf, i - 1), 0)
          beta_init[i, 1:i] = rtmvnorm(n = 1, mean = rep(0, i), sigma = diag(i),
                                   lower = lower.vec)
      }
  }
  for(i in (q+1):d) {
      beta_init[i,] = rmvnorm(n = 1, mean = rep(0, q), sigma = diag(q))
  }
  
  Sigma_init = rep(1, d)
  
  return(list(beta = beta_init, sigma = Sigma_init))
}

# Function that computes ratio corresponding to the singular values 
# d - Dimension of the data
# q0 - True no of factor
# q - Assumed no of factor
SingularRatio <- function(obj, d, q0, q) {
  net = length(obj$beta)
  singratio = rep(0, net)
  for (i in 1:net) {
    beta = matrix(unlist(obj$beta[i]), d, q)
    sing = svd(beta)$d
    singratio[i] = sum(sing[(q0+1):q])/sum(sing[1:q])
  }
  return(singratio)
}

```


```{r, echo=FALSE, message = FALSE}
#set.seed(214)
d  = 100      # number of columns in y, number of rows in beta
q0 = 10       # number of columns in beta (number of factors)
n  = 1000     # number of rows in y (number of observations)
q = q0 + 1   # number of factors we have assumed
flag = 0

thin    = 20
burnin  = 2000
seed    = 123
No.Iter = 12000    # number of iterations to run Gibbs sampler
  
MCMCRatio <- function(n, d, q0, flag = 1, thin = 20, burnin = 2000, No.Iter = 6000) {
  # Fixing some nuisance parameters
  nu = 2.2
  C0 = 1
  s = sqrt(.1/2.2)
 
  # Generating samples 
  y = GenerateData(n, d, q0, flag = 0)
  
  result = NULL
  for(i in 1:3) {
    q = q0 + i
    
    # Initializing the parameters
    beta_init = InitParam(d, q)$beta
    Sigma_init = InitParam(d, q)$sigma
    
    ## this gives us a net: No.Iter/thin posterior samples
    ## Gibbs sampler for the regular dataset: y
    obj = factGibbsMod(y = y, k = q, beta_init, Sigma_init, nu, C0, s, No.Iter, burnin, thin)
    
    # Getting ratios corresponding to the posterior samples of loading matrix after thinning 
    singratio = SingularRatio(obj, d, q0, q)
    
    summary = c(mean(singratio), var(singratio), median(singratio), quantile(singratio, names = FALSE, probs = c(0.05, 0.25, 0.75, 0.95)))
    
    result = c(result, summary)
  }
  
  return(result)
}

#MCMCRatio(n, d, q0, flag = 1, thin = 20, burnin = 2000, No.Iter = 6000)

```

```{r, echo = FALSE}
start_time <- Sys.time()

# Execute the code

nworkers <- detectCores() # Initialize the cluster 
cl <- makeCluster(nworkers)
registerDoParallel(cl)


nrep = 10

output <- foreach (rep = 1:nrep) %dopar% {
  library(mvtnorm)
  library(tmvtnorm)
  library(msm)
  library(pscl)
  library(ars)
  
  # Doing replicated simulation study
  MCMCRatio(n = 250, d = 30, q0 = 10, flag = 0, thin = 20, burnin = 2000, No.Iter = 12000)
}

stopCluster(cl) # Stop the cluster


# Get the ending time
end_time <- Sys.time()

# Calculate the elapsed time
elapsed_time <- end_time - start_time

beta = mean_beta(form_beta(obj, n, d, q))

```


```{r, echo = FALSE}
form_beta = function(obj, n, d, q) {
  beta_list = vector(mode = "list", length = n)
  for (i in 1:n) {
    beta = matrix(unlist(obj$beta[[i]]), d, q)
    beta_list[[i]] = beta
  }
  return(beta_list)
}

betalist = form_beta(obj, n, d, q)

# mean_beta = function(beta_list) {
#   beta_sum = matrix(0, d, q)
#   for (i in 1:length(beta_list)) {
#     beta_sum = beta_sum + beta_list[[i]]
#   }
#   return(beta_sum/length(beta_list))
# }

mean_beta = function(obj) {
  eigen_max = 0
  for (i in 1:length(obj$beta)) {
    eigen_max = eigen_max + svd(obj$beta[[i]])$d[1]
  }
  return(eigen_max/length(beta_list))
}


beta_eigen = mean_beta(obj)

form_sigma = function(obj, n) {
  sigma_list = vector(mode = "list", length = n)
  for (i in 1:n) {
    beta = matrix(unlist(obj$beta[i]), m, k)
    omega = diag(unlist(obj$Sigma[i]))
    sigma_list[[i]] = beta %*% t(beta) + omega
  }
  return(sigma_list)
}


form_beta_beta = function(obj, n) {
  beta_list = vector(mode = "list", length = n)
  for (i in 1:n) {
    beta = matrix(unlist(obj$beta[i]), m, k)
    beta_list[[i]] = beta %*% t(beta)
  }
  return(beta_list)
}

mean_sigma = function(sigma_list) {
  sigma_sum = matrix(0, m, m)
  for (i in 1:length(sigma_list)) {
    sigma_sum = sigma_sum + sigma_list[[i]]
  }
  return(sigma_sum/length(sigma_list))
}

mean_svd = function(A) {
  return(svd(A)$d)
}
mean_beta_svd = function(beta_list) {
  li = lapply(beta_list, mean_svd)
  return(colMeans(matrix(unlist(li), ncol = k, byrow = T)))
}



mean_beta_beta = function(beta_list) {
  beta_sum = matrix(0, m, m)
  for (i in 1:length(beta_list)) {
    beta_sum = beta_sum + beta_list[[i]]
  }
  return(beta_sum/length(beta_list))
}
```



## Comparing parameter values with posterior means

The results above seem to indicate that there is nothing wrong with the output of the posterior samples since we're able to replicate the invariant property claimed in the paper, but I think there still are some outstanding issues. 

The true value of $\beta$ compared to the posterior mean of the $\beta$'s from the Gibbs sampler do not match. 

```{r}
beta_list = form_beta(obj, length(obj$beta))
beta_est = mean_beta(beta_list)
a = svd(beta_est)$d
sum(a[(k0+1):k])/sum(a[1:k])
mean(a[(k0+1):k])/((mean(a[1:k0]) + mean(a[(k0+1):k])))
svd(W_true)$d
b = mean_beta_svd(beta_list)
sum(b[(k0+1):k])/sum(b[1:k])
```


Next, I checked if $\beta \beta'$ would match its posterior counterpart. These were particularly off.

```{r}
beta0 %*% t(beta0) # true value of beta beta'
```


```{r}
# using posterior samples of beta to estimate beta beta'
beta_beta_list = form_beta_beta(obj, length(obj$beta))
mean_beta_beta(beta_beta_list)
```


Finally, I checked if the $\Sigma$ reconstructed from the posterior samples could even accurately produce an estimate for the true $\Sigma$. They are similar, but definitely not anywhere close to perfect on the off-diagonals. 

```{r}
# true Sigma
beta0 %*% t(beta0) + omega0
```


```{r}
mean_sigma(sigma_list) # posterior estimate of Sigma
```

We can compare this to the sample covariance matrix to see how good the covariance estimation is. 

```{r}
# sample covariance matrix
1/n * t(y) %*% y
```

```{r}
# sample covariance matrix
cov(y) # same as above cuz y has 0 mean
```